Clearance Project - 3 - TKstats

### 2025-04-02 data-raw/ClearanceProject-2-RelativeLikelihood.Rmd
Updated relative loglikelihood analysis based on predictions from latest httk-dev and included best fit empirical compartmental model in the calculation.

# AIC and relative likelihood
# LL from pre-calculated httk predictions
restrictive_ll <- rph_httk_preds %>% 
  select(!httk_Preds_nonrestrictive) %>%
  rename(Conc_est = httk_Preds_restrictive,
         Time_trans = Time) %>%
  distinct() %>%
  group_by(Chemical, Species) %>%
  nest(.key = "observations") %>%
  inner_join(coef(this_pk) %>% semi_join(this_winmodel) %>% distinct()) %>%
  rowwise() %>%
  mutate(log_likelihood = log_likelihood(
    par = coefs_vector,
    data = observations,
    data_sigma_group = observations$data_sigma_group,
    dose_norm = FALSE,
    log10_trans = FALSE,
    negative = FALSE,
    includes_preds = TRUE,
    detect
  )) %>%
  select(!c(observations, coefs_vector, Time.Units, Time_trans.Units)) %>%
  mutate(model = "model_3comp2_restrictive",
         AIC = -2 * log_likelihood)

nonrestrictive_ll <- rph_httk_preds %>% 
  select(!httk_Preds_restrictive) %>%
  rename(Conc_est = httk_Preds_nonrestrictive,
         Time_trans = Time) %>%
  distinct() %>%
  group_by(Chemical, Species) %>%
  nest(.key = "observations") %>%
  inner_join(coef(this_pk) %>% semi_join(this_winmodel) %>% distinct()) %>%
  rowwise() %>%
  mutate(log_likelihood = log_likelihood(
    par = coefs_vector,
    data = observations,
    data_sigma_group = observations$data_sigma_group,
    dose_norm = FALSE,
    log10_trans = FALSE,
    negative = FALSE,
    includes_preds = TRUE,
    detect
  )) %>%
  select(!c(observations, coefs_vector, Time.Units, Time_trans.Units)) %>%
  mutate(model = "model_3comp2_nonrestrictive",
         AIC = -2 * log_likelihood) # Simple formula when there are no optimized params

httk_AIC <- bind_rows(restrictive_ll,
                      nonrestrictive_ll,
                      this_AIC %>% filter(model == "model_flat") %>%
                        filter_targets()
                      )

httk_rel_ll <- httk_AIC %>%
  distinct(Chemical, Species, model, AIC) %>%
  pivot_wider(names_from = model, values_from = AIC) %>%
  rowwise() %>% 
  mutate(minAIC = min(c_across(starts_with("model_")), na.rm = TRUE)) %>% 
  ungroup() %>%
  mutate(across(starts_with("model_"), \(x) {
    signif(exp((minAIC - x)/2) * 100, 4)
  })) %>%
  mutate(across(starts_with("model_"), \(x) {
    ifelse(x < 1E-6, 0, x)
  })) %>%
  mutate(ll_label = factor(
    case_when(model_flat > 95 ~ "Undetermined",
              model_3comp2_nonrestrictive > 95 &
                model_3comp2_restrictive > 95 ~ "Undetermined (non-flat)",
              model_3comp2_nonrestrictive > 95 ~ "Nonrestrictive",
              model_3comp2_restrictive > 95 ~ "Restrictive")
  ))

>>>>

# AIC and relative likelihood
# LL from pre-calculated httk predictions
this_pk <- this_pk %>% fit_sigma.pk(preds = rpr_httk_preds,
                                    pred_col = "httk_Preds_restrictive")
this_pk <- this_pk %>% fit_sigma.pk(preds = rpr_httk_preds,
                                    pred_col = "httk_Preds_nonrestrictive")

restrictive_ll <- this_pk$ext_fits.httk_Preds_restrictive$sigma_AIC

nonrestrictive_ll <- this_pk$ext_fits.httk_Preds_nonrestrictive$sigma_AIC

# Get NULL models and winning models AICs (call them empirical model)
invivoPKfit_AIC <- bind_rows(
  this_AIC %>% semi_join(this_winmodel) %>%
    mutate(model = "model_empirical"),
  this_AIC %>% filter(model == "model_flat")
) %>% distinct()

httk_AIC <- bind_rows(
  restrictive_ll %>% mutate(model = "model_3comp2_restrictive"),
  nonrestrictive_ll %>% mutate(model = "model_3comp2_nonrestrictive"),
  invivoPKfit_AIC
)

httk_rel_ll <- httk_AIC %>%
  distinct(Chemical, Species, model, AIC) %>%
  pivot_wider(names_from = model, values_from = AIC) %>%
  filter(!is.na(model_3comp2_nonrestrictive)) %>%
  rowwise() %>% 
  mutate(minAIC = min(c_across(starts_with("model_")), 
                      na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(across(starts_with("model_"), \(x) {
    signif(exp((minAIC - x)/2) * 100, 4)
  })) %>%
  mutate(across(starts_with("model_"), \(x) {
    ifelse(x < 1E-6, 0, x)
  })) %>%
  mutate(ll_label = factor(
    case_when(model_flat > 95 ~ "Undetermined",
              model_3comp2_nonrestrictive > 50 &
                model_3comp2_restrictive > 50 ~ "Undetermined (non-flat)",
              model_3comp2_nonrestrictive > 50 ~ "Nonrestrictive",
              model_3comp2_restrictive > 50 ~ "Restrictive")
  ))

httk_rel_ll %>% 
  select(Chemical, Species, starts_with("model_"), minAIC) %>%
  mutate(Chemical = forcats::fct_reorder2(Chemical,
                                         model_flat,
                                         minAIC,
                                         mean)
         ) %>%
  pivot_longer(cols = starts_with("model_"),
               names_to = "model", values_to = "relative_ll") %>%
  mutate(
    source = ifelse(grepl("_3comp2_", model), "httk", "invivoPKfit")
  ) %>%
  ggplot(aes(x = Chemical, y = relative_ll, color = model, shape = source)) +
  geom_point(size = 0.75, alpha = 0.9) +
  scale_color_manual(
    values = c('#33a02c', '#cc88ff', '#7b3294', 'orange'),
    breaks = c('model_empirical',
               'model_3comp2_restrictive',
               'model_3comp2_nonrestrictive',
               'model_flat')
  ) +
  scale_shape_manual(values = c(2, 3)) +
  labs(
    title = "Relative likelihoods",
    subtitle = "(rat parameterized with human CLint and fup values)",
    y = "More Likely Model >>>",
    color = ''
  ) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "bottom",
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "AIC_comparisons.png"),
       bg = "white",
       height = 5, width = 10
)

### 2025-04-02 data-raw/ClearanceProject-3-TKstats.Rmd
Updated relative loglikelihood analysis based on predictions from latest httk-dev and included best fit empirical compartmental model in the calculation.

```{r library-loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ctxR)
library(ggdist)
library(patchwork)
library(httk)
.onUnload <- function(libpath) {
  library.dynam.unload("httk", libpath)
}
```

@@

>>>>

```{r library-loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ctxR)
library(ggdist)
library(patchwork)
library(httk)
```

@@

Next I want to categorize each chemical based on putative hepatic extraction ratio.
This is calculated as:
$$
E_H = \frac{f_{up}\cdot CL_{int}}{Q_{hep} + f_{up}\cdot CL_{int}}
$$
```{r extraction_ratio}
rat_er <- rpr_param_df %>%
  group_by(Chemical) %>%
  reframe(fupClint = (Funbound.plasma * Clint),
          Qhep = (Qcardiacc*Qliverf)) %>%
  mutate(extraction_ratio = fupClint/(Qhep + fupClint),
         clearance_type_by_er = case_when(
           extraction_ratio >= 0.7 ~ "Nonrestrictive",
           extraction_ratio < 0.3 ~ "Restrictive",
           between(extraction_ratio, 0.3, 0.7) ~ "Ambiguous"
         )
  )

human_er <- rph_param_df %>%
  group_by(Chemical) %>%
  reframe(fupClint = (Funbound.plasma * Clint),
          Qhep = (Qcardiacc*Qliverf)) %>%
  mutate(extraction_ratio = fupClint/(Qhep + fupClint),
         clearance_type_by_er = case_when(
           extraction_ratio >= 0.7 ~ "Nonrestrictive",
           extraction_ratio < 0.3 ~ "Restrictive",
           between(extraction_ratio, 0.3, 0.7) ~ "Ambiguous"
         )
  )

combo_pretable <- inner_join(
  rat_er %>% select(Chemical,
                    "rat_er" = extraction_ratio, 
                    "rat_cl" = clearance_type_by_er),
  human_er %>% select(Chemical,
                    "human_er" = extraction_ratio, 
                    "human_cl" = clearance_type_by_er)
)
# Row, Columns
# In this case, rat is rows, human is columns
clearance_contingency_tbl <- table(combo_pretable$rat_cl, combo_pretable$human_cl)
addmargins(clearance_contingency_tbl)
